Next Generation Planning - Structuring and Sharing Environmental Drivers Data for the St. Lawrence

### 2019-04-28 eDriversMS-v2.Rmd
address comments and change Bray-Curtis dissimilarity for Manhattan distance

The St. Lawrence System is a vast and complex social-ecological system providing
a wealth of ecosystem services sustaining numerous economic sectors. Related
human pressures combine with climate driven environmental changes to produce
intricate exposure regimes of overlapping and potentially interacting
environmental drivers (*e.g.* hypoxia and fisheries) threatening ecosystems.
There is a critical need to systematically characterize the distribution,
intensity and overlap between drivers to support integrative initiatives such
as ecosystem-based management. While portals providing knowledge on individual
drivers exist, platforms collating comparable and interoperable knowledges
<!-- this is the first time I've ever seen the plural of knowledge... technically correct, but seems weird!--> on
multiple drivers remain conspicuously missing. This paper presents two distinct,
but related objectives. The first is to characterize the distribution and
intensity of drivers in the St. Lawrence System. The second is to launch
*eDrivers*, an open knowledge platform gathering experts committed to
structuring, standardizing and sharing knowledges on drivers in support of
science and management. We gathered data on 22 coastal, climate, fisheries and
marine traffic drivers through collaborations, existing environmental
initiatives and open data portals. We show that few areas of the St. Lawrence
are free of cumulative exposure. The Estuary, the Anticosti Gyre and coastal
areas are particularly exposed, especially in the vicinity of urban centers.
We identified 6 areas of distinct cumulative exposure regime that show that
certain drivers typically co-occur in different regions of the St. Lawrence and
that coastal areas are exposed to all driver types. Of particular concern are
two threat complexes capturing most exposure hotspots that show the convergence
of contrasting exposure regimes at the head of the Laurentian Channel. These
observations are destined to improve as *eDrivers* evolves through time to
address knowledge gaps and refine current driver layers. In an effort to share
the knowledge acquired and to ensure the lasting relevance of the description
of drivers presented in this manuscript, *eDrivers* was built on a series of
guiding principles upholding existing data management and open science
standards. Ultimately, we believe that *eDrivers* represents a much needed
solution that could radically influence broad scale research and management
practices by increasing data accessibility and interoperability and by
increasing research and decision-making efficiency.

@@

@@

Threat complexes were identified using a partional *k-medoids* clustering
algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
designed for large datasets. The CLARA algorithm uses the PAM (Partition Around
Medoids) algorithm on a sample from the original dataset to identify a set of
$k$ objects that are representative of all other objects, *i.e.* medoids and
that are central to the cluster they represent. The goal of the algorithm is to
iteratively minimize intra-cluster dissimilarity. Iterations are compared on the
basis of the average dissimilarity between cluster objects and representative
medoid to select the optimal set of $k$ medoids that minimizes average
dissimilarity. We used 100 iterations using samples of 10000 observations
(*i.e.* ~5% of observations) to identify clusters. Analyses were performed
using the *cluster* R package [@maechler2018].

@@

The difference between clusters was explored by measuring the total inter-cluster
dissimilarity and the contribution of each driver to the total inter-cluster
dissimilarity using a similarity percentage analysis (SIMPER) with Bray-Curtis
dissimilarity [Figure S4; @clarke1993]. As the drivers dataset is too
large, we used a bootstrap procedure for the SIMPER analysis, randomly selecting
5% of each cluster to run the analysis and repeating the process over 300
iterations. We also compared the mean intensity of each driver within each
cluster to better capture the inter-cluster dissimilarity. Analyses were
performed using the *vegan* R package [@oksanen2018].

@@

Intra-cluster similarity was evaluated using the Bray-Curtis similarity index
(Figure S5). As with the inter-cluster dissimilarity, we used a
bootstrap procedure for the intra-cluster similarity, randomly selecting 5% of
each cluster observation to run the analysis and repeating the process over 300
iterations. We however did not use the bootstraping procedure for clusters with
less than 10000 observations since computation time was manageable.

@@

Apart from the northeastern Gulf, the cumulative footprint of drivers is
ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
generally highest along the coast (Figure \ref{footprint}), with hotposts
located in the vicinity of coastal cities (Figure \ref{hotspot})<!-- this figure isn't loaded at the bottom -->. In general,
offshore areas are less exposed to cumulative drivers, with the Estuary and the
Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
\ref{hotspot}). This is not to say that offshore areas are free of exposure,
as most of the St. Lawrence is exposed to multiple overlapping drivers
(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
high cumulative footprint observed at the heads of the Anticosti and Esquiman
Channels (Figure \ref{footprint}).

@@

Sharing the knowledge acquired through the description of drivers in the
St. Lawrence quickly emerged as a priority to curtail the need to reach dozens
of experts across multiple organizations and over extensive periods of time to
assemble the data needed to apply integrated research and management. It is also
a requirement to ensure that this manuscript will not become a quickly outdated
snapshot of drivers distribution and intensity in the St. Lawrence System,
but rather serve as a stepping stone towards an adaptive and ever-improving
collection of knowledges.

@@

\begin{figure}[H]
\centering
\includegraphics{./Figures/footprint.png}
\caption{Distribution of driver footprint in the Estuary and Gulf of St. Lawrence. The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$. Driver layers were also log-transformed when their frequency distribution was non-normal. All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values.}
\label{footprint}
\end{figure}

>>>>

The St. Lawrence System is a vast and complex social-ecological system providing
a wealth of ecosystem services sustaining numerous economic sectors. Related
human pressures combine with climate driven environmental changes to produce
intricate exposure regimes of overlapping and potentially interacting
environmental drivers (*e.g.* hypoxia and fisheries) threatening ecosystems.
There is a critical need to systematically characterize the distribution,
intensity and overlap between drivers to support integrative initiatives such
as ecosystem-based management. While portals providing knowledge on individual
drivers exist, platforms collating comparable and interoperable knowledge on
multiple drivers remain conspicuously missing. This paper presents two distinct,
but related objectives. The first is to characterize the distribution and
intensity of drivers in the St. Lawrence System. The second is to launch
*eDrivers*, an open knowledge platform gathering experts committed to
structuring, standardizing and sharing knowledge on drivers in support of
science and management. We gathered data on 22 coastal, climate, fisheries and
marine traffic drivers through collaborations, existing environmental
initiatives and open data portals. We show that few areas of the St. Lawrence
are free of cumulative exposure. The Estuary, the Anticosti Gyre and coastal
areas are particularly exposed, especially in the vicinity of urban centers.
We identified 6 areas of distinct cumulative exposure regime that show that
certain drivers typically co-occur in different regions of the St. Lawrence and
that coastal areas are exposed to all driver types. Of particular concern are
two threat complexes capturing most exposure hotspots that show the convergence
of contrasting exposure regimes at the head of the Laurentian Channel. These
observations are destined to improve as *eDrivers* evolves through time to
address knowledge gaps and refine current driver layers. In an effort to share
the knowledge acquired and to ensure the lasting relevance of the description
of drivers presented in this manuscript, *eDrivers* was built on a series of
guiding principles upholding existing data management and open science
standards. Ultimately, we believe that *eDrivers* represents a much needed
solution that could radically influence broad scale research and management
practices by increasing data accessibility and interoperability and by
increasing research and decision-making efficiency.

@@

<!-- AQUACULTURE:
QC: MAPAQ (2016) Description des sites maricoles de la province de Québec. Sous-ministériat aux pêches et à l’aquaculture commerciales, Direction régionale des Îles-de-la-Madeleine. Ministère de l'Agriculture, des Pêcheries et de l'Alimentation du Québec. Québec, Canada
PEI: DFO (2016) PEI Aquaculture Leases. Aquaculture Division. Department of Fisheries and Oceans Canada. Prince-Edward-Island, Canada. Available at: http://www.arcgis.com/home/item.html?id=16aa8830c7084a8a92ce066b525978b4
NB: Department of Agriculture, Aquaculture and Fisheries (2016) Marine Aquaculture Site Mapping Program (MASMP). Government of New-Brunswick, New-Brunswick, Canada. Available at: https://www2.gnb.ca/content/gnb/en/departments/10/aquaculture/content/masmp.html
NS: Department of Fisheries and Aquaculture (2016) Aquaculture Site Mapping Tool. Department of Agriculture, Government of Nova-Scotia. Nova-Scotia, Canada. Available at: https://novascotia.ca/fish/aquaculture/site-mapping-tool/
NL: Department of Fisheries, Forestry and Agrifoods (2016) Description of aquaculture sites in Newfoundland. Government of Newfoundland and Labrador. Newfoundland, Canada.
-->

@@

Threat complexes were identified using a partional *k-medoids* clustering
algorithm, CLARA [CLustering for Large Applications; @kaufman1990], which was
designed for large datasets. The CLARA algorithm uses the PAM (Partition Around
Medoids) algorithm on a sample from the original dataset to identify a set of
$k$ objects that are representative of all other objects, *i.e.* medoids and
that are central to the cluster they represent. The goal of the algorithm is to
iteratively minimize intra-cluster dissimilarity. Iterations are compared on the
basis of the average dissimilarity between cluster objects and representative
medoid to select the optimal set of $k$ medoids that minimizes average
dissimilarity. We used the clustering algorithm with the manhattan distance
since this measure is less affected by extreme values [@ref]. <!-- find reference, legendre et legendre? -->
We used 100 iterations using samples of 10000 observations
(*i.e.* ~5% of observations) to identify clusters. Analyses were performed
using the *cluster* R package [@maechler2018].

@@

The difference between clusters was explored by measuring the total inter-cluster
dissimilarity and the contribution of each driver to the total inter-cluster
dissimilarity using a similarity percentage analysis (SIMPER) with Manhattan
distance [Figure S4]. The Manhattan distance was again preferred for continuity
with the clustering analysis and to ensure that outliers did not have a strong
influence the analysis. As the drivers dataset is large (~ 250000 observations),
we used a bootstrap procedure for the SIMPER analysis, randomly selecting
5% of each cluster to run the analysis and repeating the process over 300
iterations. We also compared the mean intensity of each driver within each
cluster to better capture the inter-cluster dissimilarity.

@@

Intra-cluster similarity was evaluated calculating the intra-cluster Manhattan
distance and by transforming the mean contribution to distance ($M_c$) of each
driver by $.1 / (.1 + M_c)$ to obtain a similarity measure for each driver
($S_d$). The total similarity is the sum of all $S_d$. (Figure S5).
As with the inter-cluster dissimilarity, we used a bootstrap procedure for the
intra-cluster similarity, randomly selecting 25% of each cluster observation to
run the analysis and repeating the process over 50 iterations. We only used the
bootstraping procedure for clusters with less than 40000 observations since
computation time was manageable.

@@

Apart from the northeastern Gulf, the cumulative footprint of drivers is
ubiquitous in the St. Lawrence (Figure \ref{footprint}). Cumulative exposure is
generally highest along the coast (Figure \ref{footprint}), with hotposts
located in the vicinity of coastal cities (Figure \ref{hotspot}). In general,
offshore areas are less exposed to cumulative drivers, with the Estuary and the
Anticosti Gyre being notable exceptions (Figures \ref{footprint} and
\ref{hotspot}). This is not to say that offshore areas are free of exposure,
as most of the St. Lawrence is exposed to multiple overlapping drivers
(Figures \ref{footprint} and \ref{hotspot}). For example, it is worthy to note
high cumulative footprint observed at the heads of the Anticosti and Esquiman
Channels (Figure \ref{footprint}).

@@

Sharing the knowledge acquired through the description of drivers in the
St. Lawrence quickly emerged as a priority to curtail the need to reach dozens
of experts across multiple organizations and over extensive periods of time to
assemble the data needed to apply integrated research and management. It is also
a requirement to ensure that this manuscript will not become a quickly outdated
snapshot of drivers distribution and intensity in the St. Lawrence System,
but rather serve as a stepping stone towards an adaptive and ever-improving
collection of knowledge.

@@

\begin{figure}[H]
\centering
\includegraphics{./Figures/footprint.png}
\caption{Distribution of cumulative footprint in the Estuary and Gulf of St. Lawrence. The footprint is measured by summing the relative intensity of all drivers for each grid cell: $F_d = \sum_{i=1}^{n} D_i$. Driver layers were also log-transformed when their frequency distribution was non-normal. All driver layers were also normalized between 0 and 1 using the 99th quantile to allow for direct comparison of relative intensities between drivers and control for extreme values.}
\label{footprint}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics{./Figures/hotspot.png}
\caption{Distribution of cumulative hotspots in the Estuary and Gulf of St. Lawrence. Hotspots are defined as the number of drivers in each grid cell with scaled intensity contained over their respective 80th percentile.}
\label{hotspot}
\end{figure}
